\section{Preparation}

\noindent  
This chapter covers the following ideas. When you create your lesson plan, it should contain examples which illustrate these key ideas. Before you take the quiz on this unit, meet with another student out of class and teach each other from the examples on your lesson plan. 

\input{05-Homogeneous-ODEs/Homogeneous-ODEs-objectives}


Here are the preparation problems for this unit.  All of these problems come from Schaum's Outlines.
\begin{center}
\begin{tabular}{ll}
&Preparation Problems\\
\hline\hline
Day 1& 8.33-35, 
9.1,
9.7,
9.12
\\\hline
Day 2& 10.12, 
%10.37, 
13.9, 
14.2, 
14.5
\\\hline
Day 3& 21.54,
22.36,
24.26,
8.18
%8.58
\\\hline
Day 4& Lesson Plan, Quiz
\\\hline
\end{tabular}
\end{center}


Here are the homework problems which line up with the material we are learning. If you are struggling with a topic from the preparation problem set, please use this list as a guideline to find related practice problems.


\begin{center}
\begin{tabular}{|l|c|l|l|l|l|}
\hline
Concept&Sec&Suggestions&Relevant Problems\\ \hline
Vocabulary of ODEs&8*&33-35&1-3,33-35\\ \hline
2nd Order Homogeneous&9*&1,7,12,21,27,40&1-15, 17-45\\ \hline
nth Order Homogeneous&10*&3,7,8,9,12,18,37,41,44,49&All\\ \hline
IVPs (Homogeneous)&13&9&4,9,13\\ \hline
Applications&14&2,3,5,29,31,34,41-43&1-8,26-43\\ \hline
Laplace Transforms&21*&26, 54&14(c),15(b),25,26,54-58,\\ \hline
Inverse Transforms&22*&7, 34-36,38,read 12 and 18,44&6-10,15-19,29-30,32-53\\ \hline
Solving ODES&24&26,44&5,26,31,36,43,44\\ \hline
Wronskian and Theory&8*&9,10,18,20,43,48,53,58&5-10, 13-20, 31,36-64\\ \hline
\end{tabular}
\end{center}

*The problems in these sections are quick problems. It is important to do lots of them to learn the pattern used to solve ODEs. You may be able to finish 7 problem in under 30 minutes or less.  Please do more, so that when you encounter these kinds of problems in the future you can immediately give an answer and move forward. You'll notice that the suggested list is more than 7 per day (that's because there are tons of really quick problems you can do).

